Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2023-03-14 writing/SuB-abstract/SuB-abstract.tex
abstract draft update

% 1
\vspace{-.7\baselineskip}
\noindent 
	We present experimental data that the projection of the content of attitude complements \textbf{(i)} varies between entailment-canceling operators, and \textbf{(ii)} that this by-operator variation differs between attitude predicates.
	%
	The observed variability is not captured by existing theoretical accounts of projection
	(e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}), but suggests that an analysis must consider interactions between predicates and operators, which cannot be fully determined by lexical classes. Instead, our results reveal a more nuanced picture of lexical semantic and pragmatic properties, which raises important questions for future research on projection.
	%

% 2
\noindent {\bf Projection across entailment-cancelling operators.}
	Language users may infer that a speaker who utters an attitude ascription, as in \ref{ex:family}, is committed to the content of the complement (CC),
	even when it occurs under an entailment-canceling operator, like negation \ref{ex:neg}, polar questions \ref{ex:q}, modals \ref{ex:mod}, or conditionals \ref{ex:cond}, in which case we say that it \emph{projects}.

@@

	\vspace{-.5\baselineskip}
	\noindent  Current research only rarely examines projectivity differences across these operators, with conflicting findings.
	%
	\citet{karttunen_observations_1971} proposed a distinction between English factive predicates (e.g., \emph{be annoyed, regret}), where the CC projects across all four operators, and semi-factives (e.g., \emph{discover, realize, see, notice}) where it only projects across negation.
	%
	Comparing different types of contents, \citet{smith_relationship_2014} found that that the projective content of epithets and the CC of \emph{know} was more projective under negation than conditionals,
	whereas the content of non-restrictive relative clauses and the preparatory content of \emph{win} showed the opposite pattern.
	% whereas the projective contents of non-restrictive relative clauses and \emph{win} showed the opposite pattern.
	%
	Further, \posscite{sieker_projective_2022} investigation of the projection of the CC of German purported factive and semi-factive predicates replicated \posscite{smith_relationship_2014} result for the German counterpart of `know' (\emph{wissen}), but as part of a pattern of overall higher projectivity under negation than other operators. They found no interaction with predicate type, or evidence for the (semi-)factive distinction.

	All of these studies suggest a comparatively high projectivity across negation, but the results diverge about whether they found interactions between operators and different types of contents. To investigate if the different results are due to variation between English and German, differences in contents being tested, or methodological differences, we conducted a series of experiments designed to investigate projection from under the four entailment-canceling operators in \ref{ex:family}, using the same projection measure as \citet{sieker_projective_2022} (the `certain that' diagnostic; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied it to the CC of 20 English clause-embedding predicates, including factive predicates (\emph{be annoyed, know, reveal}) and semi-factives (\emph{discover, see}),
	as well as 15 non-factive predicates, due to recent findings that their complements are also projective, albeit to varying degrees (\citealt{degen_are_2022}). Our study aimed to replicate the projection variability across operators of various English projective contents reported in Smith and Hall (2014) for English attitude complements.

@@

\noindent
{\bf Results and Analysis.} 
	\textbf{Figure~\ref{fig:figure1}} shows mean projection ratings for the 20 attitude predicates by operator; predicates are ordered by their mean rating across all operators (\emph{be annoyed} has the highest overall mean).
	%
	We find \textbf{(i)} by-operator differences in projectivity, and \textbf{(ii)} differences in by-operator variability between predicates.
	%
	The generalization \textbf{(i)} is supported by significant main effects of operator in the linear mixed effects models shown in \textbf{Table\ \ref{t:models}}. \textbf{(ii)} is supported by the fact that there are at leats $34$ significant interaction terms in these models (out of $57$ possible interactions of \texttt{operator} and \texttt{predicate}). Specifically, the models in \textbf{Table\ \ref{t:models}} illustrate specific differences across the predicates in by-operator variation with different reference levels for \texttt{predicate}: For instance, whereas the CC of \emph{be annoyed} projects more from under negation (and questions) than conditionals and modals, the CC of \emph{discover} projects less from negation than conditionals and questions, and more from under negation than modals, and the CC of \emph{know} projects less from under negation than questions, but more from under negation than modals, while the difference between negation and conditionals is not significant here.
	%
	Therefore, we did not replicate the result from \citet{smith_relationship_2014}, that the CC of \emph{know} projects more from negation than conditionals. Given that our findings indicate many interaction effects of \texttt{operator} and \texttt{predicate} throughout, our results also differ from \posscite{sieker_projective_2022} findings for German, where no such interactions were found.
	%
	Further, the variability we found is gradient, confirming previous research that by-predicate variability is not determined by categorical lexical classes (\citealt{tonhauser_how_2018,degen_are_2022}), and extending this result to by-predicate variability in interaction with entailment-cancelling contexts.

	%

@@

\noindent
{\bf Discussion:  Implications for projection analyses.}
	Our results---that projection is modulated by entailment-canceling operators and that there is by-predicate variation in the effect of operator on projection---are not captured by contemporary projection analyses, for several reasons.
	%
	The first reason is that contemporary analyses do not lead us to expect interactions with different types of entailment-cancelling operators. In \citealt{heim_projection_1983}, for instance, the CC of (semi-)factive predicates projects to the global context, except when that would produce an inconsistency, in which case the CC is accommodated to the local context of the operator. While it is conceivable for the meaning of the operator to systematically interact with the possibility of local accommodation, no such interaction has been spelled out.
	%
	The second reason is that many contemporary analyses do not make predictions for the projection of the CC of many of the 20 predicates, as they are limited to (semi-)factive predicates (e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992}, whose CCs are analyzed as presuppositions) and or entailed CCs that project unless at-issue with respect to the Question Under Discussion (e.g., \citealt{abrusan_predicting_2011,simons_best_2017}). A possible exception is the analysis of \citealt{schlenker_triggering_2021}, which predicts the potential for projection for CCs that are contextually entailed. In the full talk, we discuss how this analysis might be able to capture the gradient projection observed in our experiment.
	%
	The third reason is that contemporary projection analyses do not make sufficiently fine-grained distinctions between different clause-embedding predicates (but only between whether the CC is a presupposition or entailed). Consequently, they do not make predictions about the by-predicate variation in the effect of operator on projection.

\noindent {\bf Discussion:  Empirical implications and lexical properties.}
	%
	% {\bf THIS IS OLD, FROM ABOVE: So far, there has been no experimental investigation of by-operator variation comparing factive and semi-factive predicates. However, \citet{djarv_cognitive_2018} and \citet{tonhauser_how_2018} observed by-predicate variation in polar questions. \citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, found higher ratings for \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). Here, it is also not obvious how exactly this task relates to projection. \citet{tonhauser_how_2018} measured projection of the CC of a broad range of attitude predicates more directly, collecting ratings about speaker certainty about the CC. The observed differences did not match the expectations from Karttunen's classification (e.g., the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}).}
	%
	% \noindent
	% {\bfseries LH: } Here, JT said before that we should discuss cross-linguistic implications re Sieker \& Solstad. What would our conlusion be here?.\\
	%
	The predicate / operator interactions we find for the selected English attitude predicates are not in line with \cite{sieker_projective_2022}, who find no such effects for the selected German attitudes, but used the same task as the study presented here. This difference in findings indicates potential cross-linguistic differences in projection variability, which provides an avenue for future research.

	Our findings contradict \posscite{karttunen_observations_1971} distinction between factive and semi-factive predicates: The CC of (factive) \emph{be annoyed} does not project invariably from all four operators, and the CC \emph{discover,}  which is considered semi-factive, does not project more from under negation than the other three operators. The pattern we observed for {\em know} does not fit into either category. Similar considerations apply to \emph{see} and \emph{reveal}, which are also categorized as (semi-)factive in the literature, thus questioning the defined difference between factive and semi-factive predicates (see also \citealp{beaver_have_2010}). Future research appealing to these categories must clarify their definition. Additionally, claims about projection variability must be relativized to the entailment-canceling operator. Although our data replicate the result from \citet{tonhauser_how_2018} that, in polar questions, the CC of \emph{discover} is less projective than that of \emph{know}, the same does not hold in conditionals. Finally, our results provide further support (from negation, modals, and conditionals) for the result of \citet{degen_are_2022}, that projection does not categorically differentiate between (semi-)factive and non-factive predicates: The CCs of \emph{inform} and \emph{acknowledge}, for instance, are at least as projective as that of some (semi-)factive predicates.

@@

	So, can the observed interaction between predicate and operator in mean projection ratings be predicted from lexical semantic/pragmatic properties of the predicates, and, if so, how?
	This is a pressing question for future research, to which our data offer some tentative answers. We identify four major patterns.
	%
	The predicates \emph{pretend} and \emph{think} exhibit the {\bf `Negation high'} pattern, shown in panel (a) of \textbf{Figure~2}: We tentatively hypothesize that negation (but not the other operators) interacts with the semantic or pragmatic antiveridicality associated with these predicates.
	%
	The inferential predicates \emph{prove, confirm}, and \emph{establish} exhibit a {\bf `Negation low'} pattern, shown in panel (b): Here, we tentatively hypothesize that the veridical meaning component interacts with negation (but not the other operators), to result in lower projection ratings under negation.
	%
	For \emph{announce, confess, admit}, and \emph{reveal}, the CC is most projective when embedded in conditional antecedents: This {\bf `Conditional high'} pattern (c) may suggest that the discourse effect of a conditional interacts with the change-of-state communication predicates.
	%
	Finally, the predicates  \emph{inform, know}, and \emph{be annoyed} exhibit a {\bf `Modal low'} pattern (d). The lexical meaning of these predicates, whose CCs are among the most projective, appears to interact with the modal adverb {\em perhaps}, yielding lower projection ratings.

	% In spite of a lack of categorical distinctions about the projection behavior of our verbs, we can find some interesting patterns. \textbf{Fig.~\ref{fig:figure2}} gives the mean certainty ratings for the four operators by predicate, identifying some groups of predicates that show similar by-operator variation. We highlight four \lq projection-profiles\rq here: \emph{pretend} and \emph{think} are the only predicates that are most projective under negation compared to all other operators (\texttt{N} $>$ \texttt{M, Q, C}). \emph{annouce, confess, admit,} and \emph{reveal} are most projective under conditionals, while there is also a tendency that there is more projection from questions compared to modals and negation, a difference that may not be robust for \emph{announce} (\texttt{C} $>$ \texttt{Q} \textcolor{gray!40}{$>_?$} \texttt{M, N}). \emph{prove, confirm,} and \emph{establish} are more projective under modals and conditionals than under questions and negation (\texttt{M, C} $>$ \texttt{Q, N}). Finally, \emph{inform} and \emph{know} are most projective under questions, and least projective under modals (\texttt{Q} $>$ \texttt{N, Q} $>$ \texttt{M}).

>>>>

% 1
\vspace{-.7\baselineskip}
\noindent 
	We present experimental data that the projection of the content of attitude complements \textbf{(i)} varies between entailment-canceling operators, and \textbf{(ii)} that this by-operator variation differs between attitude predicates.
	%
	The observed variability is not captured by existing theoretical accounts of projection
	(e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}), and suggests that an analysis must consider interactions between predicates and operators, which cannot be fully determined by lexical classes. Instead, our results reveal a more nuanced picture of lexical semantic and pragmatic properties, which raises important questions for future research on projection.
	%

% 2
\noindent {\bf Projection across entailment-cancelling operators.}
	Language users may infer that a speaker who utters an attitude ascription, as in \ref{ex:family}, is committed to the content of the complement (CC, here: \emph{Julian dances Salsa.}),
	even when it occurs under an entailment-canceling operator, like negation \ref{ex:neg}, polar questions \ref{ex:q}, modals \ref{ex:mod}, or conditionals \ref{ex:cond}, in which case we say that it \emph{projects}.

@@

	\vspace{-.5\baselineskip}
	\noindent  Current research only rarely examines projectivity differences across these operators, with conflicting findings.
	%
	\citet{karttunen_observations_1971} proposed a distinction between English factive predicates (e.g., \emph{be annoyed, regret}), where the CC projects across all four operators, and semi-factives (e.g., \emph{discover, realize, see, notice}) where it only projects across negation.
	%
	Comparing different types of contents, \citet{smith_relationship_2014} found that the projective content of epithets and the CC of \emph{know} is more projective under negation than conditionals,
	whereas that of non-restrictive relative clauses and \emph{win} shows the opposite pattern.
	%
	\posscite{sieker_projective_2022} investigation of German purported factive and semi-factive predicates replicated \posscite{smith_relationship_2014} result for the German \emph{wissen} (`know'), but as part of an overall pattern of higher projectivity across negation than other operators. They found no interaction with predicate type, or evidence for the (semi-)factive distinction.

	These studies all suggest comparatively high projectivity across negation, but the results diverge about interactions between operators and different types of contents. To investigate if this is due to variation between English and German, task-differences, or different contents being tested, we conducted a series of experiments designed to investigate projection across the four entailment-canceling operators in \ref{ex:family}, using the same projection measure as \citet{sieker_projective_2022} (the `certain that' diagnostic; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied it to the CC of 20 English clause-embedding predicates, including factive predicates (\emph{be annoyed, know, reveal}) and semi-factives (\emph{discover, see}),
	as well as 15 non-factive predicates, due to recent findings that their complements are also projective, albeit to varying degrees (\citealt{degen_are_2022}). Our study aimed to replicate the projection variability across operators of various English projective contents reported in Smith and Hall (2014) for English attitude complements.

@@

\noindent
{\bf Results and Analysis.} 
	\textbf{Figure~\ref{fig:figure1}} shows mean projection ratings for the 20 attitude predicates by operator; predicates are ordered by their mean rating across all operators (\emph{be annoyed} has the highest overall mean).
	%
	Our analysis reveals two key findings: \textbf{(i)} differences in projectivity across operators, and \textbf{(ii)} differences in by-operator variability between predicates.
	%
	The first finding is supported by significant main effects of \texttt{operator} in the linear mixed effects models shown in \textbf{Table\ \ref{t:models}}.
	The second finding is supported by the three models having at least $34$ significant interaction terms (out of $57$ possible interactions of \texttt{operator} and \texttt{predicate}).

	The models in \textbf{Table\ \ref{t:models}} also illustrate some specific differences across predicates in by-operator variation (suing different reference levels for \texttt{predicate}): For instance, the CC of \emph{be annoyed} projects more from under negation (and questions) than conditionals and modals, while the CC of \emph{discover} projects less from negation than conditionals and questions, and more from under negation than modals.
	%
	Additionally, we did not replicate the result from \citet{smith_relationship_2014,sieker_projective_2022}, that the CC of \emph{know} projects more from negation than conditionals. Instead, we found that it projects less from under negation than questions, but more from under negation than modals, while the difference between negation and conditionals is not significant here.
	%
	With many interaction effects of \texttt{operator} and \texttt{predicate} throughout, our results also differ from \citet{sieker_projective_2022}, who did not find such interactions for the German attitudes they investigated.

@@

\noindent {\bf Discussion:  Empirical implications and lexical properties.}
	%
	% {\bf THIS IS OLD, FROM ABOVE: So far, there has been no experimental investigation of by-operator variation comparing factive and semi-factive predicates. However, \citet{djarv_cognitive_2018} and \citet{tonhauser_how_2018} observed by-predicate variation in polar questions. \citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, found higher ratings for \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). Here, it is also not obvious how exactly this task relates to projection. \citet{tonhauser_how_2018} measured projection of the CC of a broad range of attitude predicates more directly, collecting ratings about speaker certainty about the CC. The observed differences did not match the expectations from Karttunen's classification (e.g., the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}).}
	%
	Since our study used the same projection diagnostic, and a similar set of contents as \cite{sieker_projective_2022}, the different results indicate potential cross-linguistic variation in projection variability, suggesting a set of questions for future research. 
	%
	However, in line with their study, our findings contradict \posscite{karttunen_observations_1971} distinction between factive and semi-factive predicates: The CC of (factive) \emph{be annoyed} does not project invariably from all four operators, and the CC \emph{discover,}  which is considered semi-factive, does not project more from under negation than the other three operators. The pattern we observed for {\em know} does not fit into either category,
	thus questioning the defined difference between factive and semi-factive predicates (see also \citealp{beaver_have_2010}).
	Future research appealing to these categories must clarify their definition. Additionally, claims about projection variability must be relativized to the entailment-canceling operator.
	% Although our data replicate the result from \citet{tonhauser_how_2018} that, in polar questions, the CC of \emph{discover} is less projective than that of \emph{know}, the same does not hold in conditionals.
	%
	Finally, our results provide further support (from negation, modals, and conditionals) for the result of \citet{degen_are_2022}, that projection does not categorically differentiate between (semi-)factive and non-factive predicates: The CCs of \emph{inform} and \emph{acknowledge}, for instance, are at least as projective as that of some (semi-)factive predicates.
	%
	% Further, the variability we found is gradient, confirming previous research that by-predicate variability is not determined by categorical lexical classes (\citealt{tonhauser_how_2018,degen_are_2022}),
	% and extending this result to by-predicate variability in interaction with entailment-cancelling contexts.
	%
	In spite of a lack of categorical distinctions about the projection behavior of our verbs, we can find some interesting initial generalizations over lexical properties, indicated in \textbf{Figure\ 2}.

@@

	% So, can the observed interaction between predicate and operator in mean projection ratings be predicted from lexical semantic/pragmatic properties of the predicates, and, if so, how?
	% This is a pressing question for future research, to which our data offer some tentative answers. We identify four major patterns.
	% %
	% The predicates \emph{pretend} and \emph{think} exhibit the {\bf `Negation high'} pattern, shown in panel (a) of \textbf{Figure~2}: We tentatively hypothesize that negation (but not the other operators) interacts with the semantic or pragmatic antiveridicality associated with these predicates.
	% %
	% The inferential predicates \emph{prove, confirm}, and \emph{establish} exhibit a {\bf `Negation low'} pattern, shown in panel (b): Here, we tentatively hypothesize that the veridical meaning component interacts with negation (but not the other operators), to result in lower projection ratings under negation.
	% %
	% For \emph{announce, confess, admit}, and \emph{reveal}, the CC is most projective when embedded in conditional antecedents: This {\bf `Conditional high'} pattern (c) may suggest that the discourse effect of a conditional interacts with the change-of-state communication predicates.
	% %
	% Finally, the predicates  \emph{inform, know}, and \emph{be annoyed} exhibit a {\bf `Modal low'} pattern (d). The lexical meaning of these predicates, whose CCs are among the most projective, appears to interact with the modal adverb {\em perhaps}, yielding lower projection ratings.

	%%%%%
	% In spite of a lack of categorical distinctions about the projection behavior of our verbs, we can find some interesting patterns. \textbf{Fig.~\ref{fig:figure2}} gives the mean certainty ratings for the four operators by predicate, identifying some groups of predicates that show similar by-operator variation. We highlight four \lq projection-profiles\rq here: \emph{pretend} and \emph{think} are the only predicates that are most projective under negation compared to all other operators (\texttt{N} $>$ \texttt{M, Q, C}). \emph{annouce, confess, admit,} and \emph{reveal} are most projective under conditionals, while there is also a tendency that there is more projection from questions compared to modals and negation, a difference that may not be robust for \emph{announce} (\texttt{C} $>$ \texttt{Q} \textcolor{gray!40}{$>_?$} \texttt{M, N}). \emph{prove, confirm,} and \emph{establish} are more projective under modals and conditionals than under questions and negation (\texttt{M, C} $>$ \texttt{Q, N}). Finally, \emph{inform} and \emph{know} are most projective under questions, and least projective under modals (\texttt{Q} $>$ \texttt{N, Q} $>$ \texttt{M}).

\noindent
{\bf Theoretical Implications.}
	Our results---that projection is modulated by entailment-canceling operators and that there is by-predicate variation in the effect of operator on projection---are not captured by contemporary projection analyses, for several reasons.
	%
	The first reason is that contemporary analyses do not lead us to expect interactions with different types of entailment-cancelling operators. In \citealt{heim_projection_1983}, for instance, the CC of (semi-)factive predicates projects to the global context, except when that would produce an inconsistency, in which case the CC is accommodated to the local context of the operator. While it is conceivable for the meaning of the operator to systematically interact with the possibility of local accommodation, no such interaction has been spelled out.
	%
	The second reason is that many contemporary analyses do not make predictions for the projection of the CC of many of the 20 predicates, as they are limited to (semi-)factive predicates (e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992}, whose CCs are analyzed as presuppositions) and or entailed CCs that project unless at-issue with respect to the Question Under Discussion (e.g., \citealt{abrusan_predicting_2011,simons_best_2017}). A possible exception is the analysis of \citealt{schlenker_triggering_2021}, which predicts the potential for projection for CCs that are contextually entailed. In the full talk, we discuss how this analysis might be able to capture the gradient projection observed in our experiment.
	%
	The third reason is that contemporary projection analyses do not make sufficiently fine-grained distinctions between different clause-embedding predicates (but only between whether the CC is a presupposition or entailed). Consequently, they do not make predictions about the by-predicate variation in the effect of operator on projection.
